{\bf Intrusion Detection through Provenance-Based Histograms}

### 2013-11-24 writeup/paper.tex
Fixed figure numbers

\subsection{Examples}
In Figure \ref{kde-example}, we see a few examples of a generated histogram and KDE model. The green bars represent the histogram for the particular statistic, and the blue curve is the KDE. Finally, the red line is the value for a ``test" node. In this case, the test node appears to be a possible exploit, because it intersects the KDE at low values. 
\begin{figure*}
  \label{kde-example}
  \caption{Example histograms of a centrality measure and process statistic for the \texttt{gcc} exploit, with the associated KDE. The green bars represent the raw histogram counts, the blue curve represents the kernel density function, and the red vertical line represents the estimation of a node of interest.}
  \centering
    \includegraphics[width=\textwidth]{img/hist.png}
\end{figure*}

@@

We ran \texttt{hello} normally 40 times on various inputs and collected the provenance times. We then exploited the buffer overflow vulnerability in \texttt{hello} to spawn an \texttt{ls} process, and saved the standard out of that process to a file. See Figure \ref{hello-orbiter} for provenance graphs displaying normal and exploited behavior of \texttt{hello}.

@@

We ran our program on over 200 good inputs, and stored the provenance data. We then ran our program on two bad inputs, first to call \texttt{ls} as a Proof of Concept, and again to call \texttt{/bin/sh} to demonstrate that more severe attacks can be performed. We again exported the provenance generated by these steps, for comparison with our good usage. This is similar to recording all uses of a program during a server's lifetime. See Figure \ref{mcrypt-orbiter} displaying normal and exploited behavior for our version of mcrypt.

@@

\subsection{Selecting Metrics}
Because we had no prior intuition into what the histograms on provenance data would actually look like, we approached selecting statistics from an objective, data-based standpoint. We wanted to use statistics that had variation over our workload to minimize the dependency of our statistics on the workload data. Towards this end, we graphed all of the statistics over the ``normal" provenance data, to see where there were trends and variation. Our results for GCC are in Figures \ref{simple-gcc} and \ref{centrality-gcc}. We can see that proc\_out\_proc (outgoing processes), the ancestor centrality, and Opsahl Closeness Centrality statistics have more variation, whereas the other statistics were unimodal or bimodal (0 or 1).

@@

\subsection{Results}
\subsection{Provenance Data}
After running the normal and exploited programs described above, we visualized the provenance graphs using Orbiter \cite{orbiter}. After searching for the nodes named after our programs, we were able to find both the clean and exploited nodes. The clean nodes for ``hello" simply read in one file through a pipe and outputted one file, whereas the exploited version clearly has a forked \texttt{/usr/bin/ls} process. Similarly, the mcrypt simulation (implemented in Python) has only an output file as a descendent normally, but when exploited has multiple extra processes/files as descendants (see Figure \ref{hello-orbiter} and Figure \ref{mcrypt-orbiter}). Similarly, the normal GCC and exploited GCC displayed very different provenance graphs, but they were too big to fit legibly on a page.

\subsection{Statistical Data and KDE Predictions}
Table \ref{results} summarizes some of the preliminary results we have generated. Here we have included, for each statistic and test, the average over the "normal" nodes, the observed value for the "test" node, and the KDE probability output for that value. We have chosen four metrics that provided a statistically significant result. 
% The other statistics failed to be as powerful (see Table 2 XXX).
% from http://hhh123.wordpress.com/2008/08/13/latex-table-across-two-columns/
% TODO: include difference/ratio from normal?
\begin{table*}[ht]
\label{results}
{\small
  \begin{center}
  \begin{tabular}{| l | l | l | l |}
    \hline
    metric & hello & mcrypt & gcc \\ \hline
     \# forked processes & 0.0 | 1 | 0.108 & 0.0 | 1 | 0.107 & 0.9 | 0 | 0.408 \\ \hline
     \# input files & 0.5 | 0 | 0.453 & 0.5 | 0 | 0.453 & 0.5 | 1 | 0.107 \\ \hline
     \# ouptut files & 0.5 | 1 | 0.453 & 0.5 | 1 | 0.453 & 0.0 | 3 | 0.000 \\ \hline
    Opsahl centrality & 1.791 | 2.83 | 0.252 & 1.250 | 4.867 | 0.000 & 2.793 | 7.415 | 0.024 \\
    \hline
  \end{tabular}
  \end{center}
}
\hfill{}
\caption{average over good | test node value | kernel density estimate of test node being ``normal"
}
\label{tb:tablename}
\end{table*}

>>>>

\subsection{Examples}
In Figure 1, we see a few examples of a generated histogram and KDE model. The green bars represent the histogram for the particular statistic, and the blue curve is the KDE. Finally, the red line is the value for a ``test" node. In this case, the test node appears to be a possible exploit, because it intersects the KDE at low values. 
\begin{figure*}
  \label{kde-example}
  \caption{Example histograms of a centrality measure and process statistic for the \texttt{gcc} exploit, with the associated KDE. The green bars represent the raw histogram counts, the blue curve represents the kernel density function, and the red vertical line represents the estimation of a node of interest.}
  \centering
    \includegraphics[width=\textwidth]{img/hist.png}
\end{figure*}

@@

We ran \texttt{hello} normally 40 times on various inputs and collected the provenance times. We then exploited the buffer overflow vulnerability in \texttt{hello} to spawn an \texttt{ls} process, and saved the standard out of that process to a file. See Figure 2 for provenance graphs displaying normal and exploited behavior of \texttt{hello}.

@@

We ran our program on over 200 good inputs, and stored the provenance data. We then ran our program on two bad inputs, first to call \texttt{ls} as a Proof of Concept, and again to call \texttt{/bin/sh} to demonstrate that more severe attacks can be performed. We again exported the provenance generated by these steps, for comparison with our good usage. This is similar to recording all uses of a program during a server's lifetime. See Figure 3 displaying normal and exploited behavior for our version of mcrypt.

@@

\subsection{Selecting Metrics}
Because we had no prior intuition into what the histograms on provenance data would actually look like, we approached selecting statistics from an objective, data-based standpoint. We wanted to use statistics that had variation over our workload to minimize the dependency of our statistics on the workload data. Towards this end, we graphed all of the statistics over the ``normal" provenance data, to see where there were trends and variation. Our results for GCC are in Figures 4 and 5. We can see that proc\_out\_proc (outgoing processes), the ancestor centrality, and Opsahl Closeness Centrality statistics have more variation, whereas the other statistics were unimodal or bimodal (0 or 1).

@@

\subsection{Results}
\subsection{Provenance Data}
After running the normal and exploited programs described above, we visualized the provenance graphs using Orbiter \cite{orbiter}. After searching for the nodes named after our programs, we were able to find both the clean and exploited nodes. The clean nodes for ``hello" simply read in one file through a pipe and outputted one file, whereas the exploited version clearly has a forked \texttt{/usr/bin/ls} process. Similarly, the mcrypt simulation (implemented in Python) has only an output file as a descendent normally, but when exploited has multiple extra processes/files as descendants (see Figure 2 and Figure 3). Similarly, the normal GCC and exploited GCC displayed very different provenance graphs, but they were too big to fit legibly on a page.

\subsection{Statistical Data and KDE Predictions}
Table 1 summarizes some of the preliminary results we have generated. Here we have included, for each statistic and test, the average over the "normal" nodes, the observed value for the "test" node, and the KDE probability output for that value. We have chosen four metrics that provided a statistically significant result. 
% The other statistics failed to be as powerful (see Table 2 XXX).
% from http://hhh123.wordpress.com/2008/08/13/latex-table-across-two-columns/
% TODO: include difference/ratio from normal?
\begin{table*}[ht]
\label{results}
{\small
  \begin{center}
  \begin{tabular}{| l | l | l | l |}
    \hline
    metric & hello & mcrypt & gcc \\ \hline
     \# forked processes & 0.0 | 1 | 0.108 & 0.0 | 1 | 0.107 & 0.9 | 0 | 0.408 \\ \hline
     \# input files & 0.5 | 0 | 0.453 & 0.5 | 0 | 0.453 & 0.5 | 1 | 0.107 \\ \hline
     \# ouptut files & 0.5 | 1 | 0.453 & 0.5 | 1 | 0.453 & 0.0 | 3 | 0.000 \\ \hline
    Opsahl centrality & 1.791 | 2.83 | 0.252 & 1.250 | 4.867 | 0.000 & 2.793 | 7.415 | 0.024 \\
    \hline
  \end{tabular}
  \end{center}
}
\hfill{}
\caption{average over good | test node value | kernel density estimate of test node being ``normal"
}
\label{tb:tablename}
\end{table*}
